---
title: "Reproducible Research - Peer Assessment 1"
output:
  html_document:
    keep_md: yes
---

## Loading and preprocessing the data

```{r DataPrep}
library(ggplot2)
library(pander)

# read the csv file
AC <- read.csv("Activity.csv")

# calculate the total steps per day
AC.sum <- aggregate(formula=steps~date, data=AC, FUN=sum, na.action = na.omit)
colnames(AC.sum)[colnames(AC.sum)=="steps"] <- "Total.Steps"

## calculate the mean steps per day
AC.mean <- aggregate(formula=steps~date, data=AC, FUN=mean, na.action = na.omit)
colnames(AC.mean)[colnames(AC.mean)=="steps"] <- "Mean.Steps"

# calculate the medians steps per day
AC.median <- aggregate(formula=steps~date, data=AC, FUN=median, na.action = na.omit)
colnames(AC.median)[colnames(AC.median)=="steps"] <- "Median.Steps"

# merge Total, Mean, Median tables together
AC.all <- merge(AC.sum, AC.mean, by="date", all=TRUE)
AC.all <- merge(AC.all, AC.median, by="date", all=TRUE)

# calculate the mean of 5 min interval across all days.
AC.5min <- aggregate(formula=steps~interval, data=AC, FUN=mean, na.action = na.omit)
colnames(AC.5min)[colnames(AC.5min)=="steps"] <- "mean.5min.steps"



```

## Total, Mean, and Median number of steps taken per day

```{r AllTable, echo=TRUE, results='asis'}
# print table
pandoc.table(AC.all, justify = "left", split.tables = Inf)
```

## Histogram of Total Number of Steps Taken Each Day

```{r Hist, echo=TRUE, results='asis'}
# create a histogram per day
g <- ggplot(AC.sum, aes(AC.sum$Total.Steps))
g <- g + geom_histogram(color="white")
g <- g + ggtitle("Histogram of Total Number of Steps Taken Each Day") 
g <- g + xlab("Total Number of Steps") + ylab("Frequency")
g <- g + theme_bw()
g <- g +  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5), title=element_text(size=9))
print(g)
rm(g) 
```

#Line Chart of average daily activity pattern per 5 min interval
Time Series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r, echo=TRUE, results='asis'}
# plot Cumulative Count per day
g <- ggplot(AC.5min, aes(AC.5min$interval, AC.5min$mean.5min.steps)) 
g <- g + geom_line(color="blue", linetype=1, size=1) 
g <- g + ggtitle("Time series plot of 5-min interval & average num steps across all days") 
g <- g + xlab("5 minute interval") 
g <- g + ylab("Average Number of Steps")
g <- g + theme_bw()
g <- g + scale_x_continuous(breaks = round(seq(min(AC.5min$interval), max(AC.5min$interval), by = 100),1))
g <- g +  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5), title=element_text(size=11)) # vertical x aXIS TEXT
print(g)
```

# Max Steps in 5 min interval
Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r, echo=TRUE, results='asis'}
max.5min <- max(AC.5min$mean.5min.steps)
print(max.5min)
```

# Imputing missing values
1.Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r, echo=TRUE, results='asis'}
AC.steps.na <- subset(AC, is.na(steps) )
AC.date.na <- subset(AC, is.na(date) )
AC.interval.na <- subset(AC, is.na(interval) )
AC.anyrow.na <- subset(AC, is.na(steps) | is.na(date) | is.na(interval))
missing.values <- data.frame(Total.number.of.rows=nrow(AC),
                            Missing.steps.values=nrow(AC.steps.na),
                            Missing.date.values=nrow(AC.date.na),
                            Missing.interval.values=nrow(AC.interval.na),
                            Rows.with.missing.values=nrow(AC.anyrow.na)
                            )
# print table
pandoc.table(t(missing.values), justify = "left", split.tables = Inf)

rm(missing.values)

```

# Imputing missing values
2.Strategy for filling in all of the missing values in the dataset. 

My strategy: replace the NAs in the dataset with the mean of the 5 min interval across all days.

```{r, echo=TRUE, results='asis'}
# merge the mean for 5min interval columns with the main data set
AC.clean <- merge(AC, AC.5min, by="interval", all=TRUE)
AC.clean$steps[is.na(AC.clean$steps)] <- AC.clean$mean.5min.steps[is.na(AC.clean$steps)]

# print again to make sure that no NAs are left in the data set
AC.clean.steps.na <- subset(AC.clean, is.na(steps) )
AC.clean.date.na <- subset(AC.clean, is.na(date) )
AC.clean.interval.na <- subset(AC.clean, is.na(interval) )
AC.clean.anyrow.na <- subset(AC.clean, is.na(steps) | is.na(date) | is.na(interval))
missing.values <- data.frame(Total.number.of.rows=nrow(AC.clean),
                            Missing.steps.values=nrow(AC.clean.steps.na),
                            Missing.date.values=nrow(AC.clean.date.na),
                            Missing.interval.values=nrow(AC.clean.interval.na),
                            Rows.with.missing.values=nrow(AC.clean.anyrow.na)
                            )
# print table
pandoc.table(t(missing.values), justify = "left", split.tables = Inf)

```

## Total, Mean, and Median number of steps taken per day for the CLEAN data set compared to the ORIGINAL data set
Notice that the clean data set does not have any NAs, and that the NA values in the original data set are now replaced with actual values in the clean data set.

```{r DataPrepClean, echo=TRUE, results='asis'}
# calculate the total steps per day
AC.clean.sum <- aggregate(formula=steps~date, data=AC.clean, FUN=sum, na.action = na.omit)
colnames(AC.clean.sum)[colnames(AC.clean.sum)=="steps"] <- "clean.Total.Steps"

## calculate the mean steps per day
AC.clean.mean <- aggregate(formula=steps~date, data=AC.clean, FUN=mean, na.action = na.omit)
colnames(AC.clean.mean)[colnames(AC.clean.mean)=="steps"] <- "clean.Mean.Steps"

# calculate the medians steps per day
AC.clean.median <- aggregate(formula=steps~date, data=AC.clean, FUN=median, na.action = na.omit)
colnames(AC.clean.median)[colnames(AC.clean.median)=="steps"] <- "clean.Median.Steps"

# merge Total, Mean, Median tables together
AC.clean.all <- merge(AC.clean.sum, AC.clean.mean, by="date", all=TRUE)
AC.clean.all <- merge(AC.clean.all, AC.clean.median, by="date", all=TRUE)

# merge with original data set to show the difference
AC.clean.all <- merge(AC.clean.all, AC.all, by=c("date"), all=TRUE)

# print the table to compare clean vs. original ordering the columns to have totals/mean/median next to each other.
pandoc.table(AC.clean.all[,c(1,2,5,3,6,4,7)], justify = "left", split.tables = Inf)
```

## Histogram of Total Number of Steps Taken Each Day for the CLEAN data set

```{r , echo=TRUE, results='asis'}
# create a histogram per day
g <- ggplot(AC.clean.sum, aes(AC.clean.sum$clean.Total.Steps))
g <- g + geom_histogram(color="white")
g <- g + ggtitle("Histogram of Total Number of Steps Taken Each Day") 
g <- g + xlab("Total Number of Steps") + ylab("Frequency")
g <- g + theme_bw()
g <- g +  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5), title=element_text(size=9))
print(g)
rm(g) 
```

## Are there differences in activity patterns between weekdays and weekends?

```{r , echo=TRUE, results='asis'}
# add the weekday or weekend flag
AC$day <- weekdays(as.Date(AC$date))
AC$dayType <- ifelse(AC$day=="Saturday" | AC$day=="Sunday", "Weekend", "Weekday")
AC$dayType <- as.factor(AC$dayType)

# calculate average steps for weekday and weeekend for each 5 min interval
AC.dayType <- aggregate(formula=steps~interval+dayType, data=AC, FUN=mean)

```

# Plot 5 min interval for Weekdays and Weekends

```{r , echo=TRUE, results='asis'}
# plot Cumulative Count per day
g <- ggplot(AC.dayType, aes(AC.dayType$interval, AC.dayType$steps)) 
g <- g + geom_line(color="blue", linetype=1, size=1) 
g <- g + facet_grid( dayType~.)
g <- g + ggtitle("Time series plot of 5-min interval & average num steps across day type") 
g <- g + xlab("5 minute interval") 
g <- g + ylab("Average Number of Steps")
g <- g + theme_bw()
g <- g + scale_x_continuous(breaks = round(seq(min(AC.5min$interval), max(AC.5min$interval), by = 100),1))
g <- g +  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5), title=element_text(size=11)) # vertical x aXIS TEXT
print(g)
```
